---
title: "Peak Matrix Processing for -omics data sets"
author: 
    name: "Andris Jankevics"
    affiliation: Phenome Centre Birmingham, University of Birmingham
    email: a.jankevics@bham.ac.uk

package: pmp

date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2  
        number_sections: true  
        toc_float: true

vignette: >
    %\VignetteIndexEntry{Peak Matrix Processing for -omics data sets}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

R package pmp (Peak Matrix Processing for -omics data sets) contains many
useful functions commonly used for -omics data pre-processing.

Offered functionality includes data filtering based on fraction of missing
values across samples of features. Blank sample filter for LC-MS of DI-MS 
data sets and feature filter based on quality control (QC)
sample RSD \% values.

In this package we have implemented simplified unified interface for several
missing value imputation algorithms. Supported methods are k-nearest neighbors 
(knn), random forests (rf), Bayesian PCA missing value estimator (bpca), 
mean or median value of the given feature and constant small value.

Variance stabilising generalised logarithm transformation (glog) algorithm is 
implimented to help to minimise contributions from unwanted technical variaton
of sample collection.

PQN and normalisation the the total signal intensity methods are implemented for
normalisation of biological variability across measured samples.

# Installation

```{r eval=FALSE, include=TRUE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("pmp")
```


# Standard peak matrix processing workflow for LC/MS or DI/MS data sets

```{r setup}
library(pmp)
```

