---
title: "Signal and batch correction for mass spectrometry"
author: 
    name: "Andris Jankevics"
    affiliation: Phenome Centre Birmingham, University of Birmingham
    email: a.jankevics@bham.ac.uk
package: pmp
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        toc_depth: 2  
        number_sections: true  
        toc_float: true

bibliography: pmp.bib

vignette: >
    %\VignetteIndexEntry{sbcms}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

This vignette demonstrates how to apply Quality Control-Robust Spline 
Correction (QC-RSC) [@kirwan2013] algorithm for signal drift and batch effect
correction within/across a multi-batch direct infusion mass spectrometry (DIMS)
and liquid chromatography mass spectrometry (LCMS) data sets.

Please read "Signal and batch correction, data assessment and 
correction" vignette to learn how to assess your data set and details on
algorithm itself.

# Installation

You should have R version 4.0.0 or above and Rstudio installed to be able to
run this notebook.

Execute following commands from the R terminal.

```{r echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
install.packages("gridExtra")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("pmp")
```

Load the required libraries into the R environment
```{r, message=FALSE, warning=FALSE}
library(S4Vectors)
library(SummarizedExperiment)
library(pmp)
library(ggplot2)
library(reshape2)
library(gridExtra)
```

# Data set

In this tutorial we will be using  an direct infusion mass spectrometry (DIMS) 
data set consisting of 172 samples measured across 8 batches and is included in
`pmp` package as `SummarizedExperiemnt` class object `MTBLS79`. 
More detailed description of the data set is available from  @kirwan2014, 
[MTBLS79](https://www.ebi.ac.uk/metabolights/MTBLS79) and R man page.

```{r, eval=FALSE}
help ("MTBLS79")
```

```{r}
data("MTBLS79")

class <- MTBLS79$Class
batch <- MTBLS79$Batch
sample_order <- c(1:ncol(MTBLS79))

# Input data structure
MTBLS79

class[1:10]
batch[1:10]
sample_order[1:10]
```

# Filtering a data set

Current implementation of `QCRSC` algorithm does support missing values in the 
input data object, but we would recommend to filter out features which were net 
reproducibly measured across quality control (QC) sample. In this example we 
will use 80% detection threshold.

```{r}
data <- filter_peaks_by_fraction(df=MTBLS79, classes=class, method="QC",
    qc_label="QC", min_frac=0.8)
```

# Applying signal and batch correction 

Function `QCRSC` should be used to apply signal batch correction. 

Argument `df` should be `SummarizedExperiment` object or matrix-like R data 
structure with all `numeric()` values.

Argument `order` should be `numeric()` vector containing sample injection 
order during analytical measurement and should be the same length as number of
sample in the input object. 

Argument `batch` should be `numeric()` or `character()` vector containing values
of sample batch identifier. If all samples were measured in 1 batch, then all 
values in the `batch` vector should be identical. 

Values for `classes` should be character vector containing sample class labels.
Class label for quality control sample has to be `QC`.

```{r}
corrected_data <- QCRSC(df=data, order=sample_order, batch=batch, 
    classes=class, spar=0, minQC=4)
```

# Visualising results

Function 'sbc_plot' provides visual comparison of the data before and after 
correction. For example we can check output for features '1', '5', and '30' in 
peak matrix.
```{r message=FALSE, warning=FALSE, fig.height=6, fig.width=6}
plots <- sbc_plot (df=MTBLS79, corrected_df=corrected_data, classes=class, 
    batch=batch, output=NULL, indexes=c(1, 5, 30))
plots
```

The scores plots of principal components analysis (PCA) before 
and after correction can be used to asses effects of data correction. 

In this example, `PQN` method is used to normalise data,
`KNN` for missing value imputation and `glog` for data scaling. All functions 
are availiable as a part of `r Biocpkg("pmp")` package.

A more detailed overview on data pre-processing is detailed in @guida2016.

```{r, fig.width=6, fig.height=8}

manual_color = c("#386cb0", "#ef3b2c", "#7fc97f", "#fdb462", "#984ea3", 
    "#a6cee3", "#778899", "#fb9a99", "#ffff33")

pca_data <- pqn_normalisation(MTBLS79, classes=class, qc_label="QC")
pca_data <- mv_imputation(pca_data, method="KNN", k=5, rowmax=0.5,
    colmax=0.5, maxp=NULL, check_df=FALSE)
pca_data <- glog_transformation(pca_data, classes=class, qc_label="QC")

pca_corrected_data <- pmp::pqn_normalisation(corrected_data, classes=class,
    qc_label="QC")
pca_corrected_data <- pmp::mv_imputation(pca_corrected_data, method="KNN", k=5,
    rowmax=0.5, colmax=0.5, maxp=NULL, check_df=FALSE)
pca_corrected_data <- pmp::glog_transformation(pca_corrected_data, 
    classes=class, qc_label="QC")

pca_data <- prcomp(t(assay(pca_data)), center=TRUE, scale=FALSE)
pca_corrected_data <- prcomp(t(assay(pca_corrected_data)), center=TRUE, scale=FALSE)

# Calculate percentage of explained variance of the first two PC's
exp_var_pca <- round(((pca_data$sdev^2)/sum(pca_data$sdev^2)*100)[1:2],2)
exp_var_pca_corrected <- round(((pca_corrected_data$sdev^2) /
    sum(pca_corrected_data$sdev^2)*100)[1:2],2)

plots <- list()

plotdata <- data.frame(PC1=pca_data$x[, 1], PC2=pca_data$x[, 2], 
    batch=as.factor(batch), class=class)

plots[[1]] <- ggplot(data=plotdata, aes(x=PC1, y=PC2, col=batch)) +
    geom_point(size=2) + theme(panel.background=element_blank()) +
    scale_color_manual(values=manual_color) +
    ggtitle("PCA scores, before correction") +
    xlab(paste0("PC1 (", exp_var_pca[1] ," %)")) +
    ylab(paste0("PC2 (", exp_var_pca[2] ," %)"))

plots[[2]] <- ggplot(data=plotdata, aes(x=PC1, y=PC2, col=class)) +
    geom_point(size=2) + theme(panel.background=element_blank()) +
    scale_color_manual(values=manual_color) +
    ggtitle("PCA scores, before correction") +
    xlab(paste0("PC1 (", exp_var_pca[1] ," %)")) +
    ylab(paste0("PC2 (", exp_var_pca[2] ," %)"))

plotdata_corr <- data.frame(PC1=pca_corrected_data$x[, 1], 
    PC2=pca_corrected_data$x[, 2], batch=as.factor(batch), class=class)

plots[[3]] <- ggplot(data=plotdata_corr, aes(x=PC1, y=PC2, col=batch)) +
    geom_point(size=2) +
    theme(panel.background=element_blank()) +
    scale_color_manual(values=manual_color) +
    ggtitle("PCA scores, after correction") +
    xlab(paste0("PC1 (", exp_var_pca_corrected[1] ," %)")) +
    ylab(paste0("PC2 (", exp_var_pca_corrected[2] ," %)"))

plots[[4]] <- ggplot(data=plotdata_corr, aes(x=PC1, y=PC2, col=class)) +
    geom_point(size=2) +
    theme(panel.background=element_blank()) +
    scale_color_manual(values=manual_color) +
    ggtitle("PCA scores, after correction") +
    xlab(paste0("PC1 (", exp_var_pca_corrected[1] ," %)")) +
    ylab(paste0("PC2 (", exp_var_pca_corrected[2] ," %)"))

grid.arrange(ncol=2, plots[[1]], plots[[2]], plots[[3]], plots[[4]])
```

# Session information

```{r}
sessionInfo()
```

# References
